\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-20 paper.tex
Yans updates

% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
% Genome ARGs &
% Event ARGs
In the following, we define the genome ARG (gARG) and contrast it with the event ARGs (eARG).
% Retrospective vs prospective
% TODO: This needs revision - see also https://github.com/tskit-dev/what-is-an-arg-paper/pull/380
We continue with a description of the usually held retrospective view of ARGs,
but note that ARGs can also be viewed as prospective, particularly in forwards-time
simulations. In the context of forwards-time simulations, we mention the
simplification operation, which removes ancestral genome material that was not
inherited.
% Converting and eARG to gARG &
% ARGs and local trees &
% Locally unary nodes &
% Levels of simplification
This simplification is closely related to operations needed to convert eARG and gARG,
which opens a rich set of details about ARGs and local trees, local unary nodes, and
levels of simplification.
% Precision of recombination information
We then highlight the important difference between the required level of precision
about recombination information for eARGs compared to gARGs.
% Computational efficiency and data interchange
The way this recombination information is stored in eARGs and gARGs has important
implications for computational efficiency, which is critical if we want to use ARGs
for modern genomic datasets.
% Discussion
We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
Finally, the literature on ARGs and associated aspects is vast, particularly
due to different definitions and implementations. We cover some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
description of the big and little ARGs (Appendix~\ref{sec-big-and-little-arg}),
survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
ARGs at an individual vs cell lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

@@

% Here we describe recombination features
A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args} A \& B, each recombination node has exactly 2 parents,
corresponding to a single recombination event with a specific breakpoint. Assuming that each
breakpoint occurs a a different location (the "infinite breakpoints" model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args} C \& D have ARG nodes with greater
than two parents, indicating deliberate imprecision: these are cases where multiple recombination events
have been ``pushed down'' onto a single more recent node.
Note, however, that the number
of breakpoints in these two ARGs is smaller than the number of ARG nodes with multiple parents,
because those nodes frequently share breakpoint positions. The sharing of breakpoint locations
between multiple-parent nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.

% Here we move on to coalescence
As for coalescence in these inferred ARGs, the shading (``coalescent span'')
reveals that many non-recombinant (green) nodes are
locally unary (see section \ref{sec-locally-unary-edges}). In other words,
in some regions of the genome these nodes are present in the ancestry without
corresponding to coalescence of ancestral material. As argued previously, this can provide useful
information about shared ancestry. Note however, that such unary segments are not inferred
in the \relate\ ARG: other than the samples it has entirely solid nodes.
Only the \tsinfer\ results have nodes with more than two children (polytomies);
in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
uncertainty in the order of coalescent events, in an equivalent
manner to the imprecision associated with recombination nodes having more
than two parents. The clearest example is the polytomy involving the nodes
\noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}, which happen to
have identical sequences. \kwarg, \argweaver, and \relate\ require
bifurcating trees, so are forced
to pick an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
relationship between these samples). 
The number of nodes in Fig.~\ref{fig-inferred-args} C
that have more than two children also hints at considerable uncertainty in
the order of branching elsewhere in the ARG.

@@

% TODO add additional explanation of the bottom row, if enough isn't written below
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
We can see that both \tsinfer and (to a lesser extent) \relate have edges
that span multiple tree boundaries, indicating that they certainly
do not infer a series of \emph{unrelated} local trees.
%(i.e., with either unlabelled internal
%nodes or differently labelled nodes in each tree).
For example, both infer that \noderef{Af-f}, \noderef{Fr-f} and \noderef{Wa-f}
share a common ancestor along
the entire sequence. They do not, however, show extensive node sharing in the
oldest time periods, reflecting the fact that they do not encode topology
above the local tree roots (i.e. they do not go back to a
"grand MRCA"). This is also seen in the fact they their ARGs contain multiple roots
(nodes with zero parents, in blue), whereas \kwarg and \argweaver have a
single ARG root.
These two methods contain a level of detail about recombination
that lies somewhere between a sequence of unrelated local trees on one
extreme and a completely precise event ARG on the other.

>>>>

% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
% Genome ARGs &
% Event ARGs
In the following, we define a genome ARG (gARG) and contrast it with event ARGs (eARGs).
% Retrospective vs prospective
% TODO: This needs revision - see also https://github.com/tskit-dev/what-is-an-arg-paper/pull/380
We continue with a description of the usually held retrospective view of ARGs,
but note that ARGs can also be viewed as prospective, particularly in forwards-time
simulations. In the context of forwards-time simulations, we mention the
simplification operation, which removes ancestral genome material that was not
inherited.
% Converting and eARG to gARG &
% ARGs and local trees &
% Locally unary nodes &
% Levels of simplification
This simplification is closely related to operations needed to convert eARG and gARG,
which opens a rich set of details about ARGs and local trees, local unary nodes, and
levels of simplification.
% Precision of recombination information
We then highlight the important difference between the required level of precision
about recombination information for eARGs compared to gARGs.
% Computational efficiency and data interchange
The way this recombination information is stored in eARGs and gARGs has important
implications for computational efficiency, which is critical if we want to use ARGs
for modern genomic datasets.
% Discussion
We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
Finally, the literature on ARGs and associated aspects is vast, particularly
due to different definitions and implementations. We cover some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
description of the big and little ARGs (Appendix~\ref{sec-big-and-little-arg}),
survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
ARGs at an individual vs cell lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

@@

% Here we describe recombination features
A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args} A \& B, each recombination event is tagged by
a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
occurs at a different location (the "infinite breakpoints" model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args} C \& D have ARG nodes with multiple
parents and multiple children. These indicate deliberate imprecision, where one
or more recombinations have been ``pushed down'' onto a more recent, identifiable node.
Several of these nodes have more than two parents (in red), representing
multiple recombinations inferred to have occurred in an unresolvable order.
Nevertheless, the number of breakpoints in these two ARGs is smaller than the
number of multiple-parent ARG nodes, showing that several multiple-parent nodes
must share breakpoint positions. The sharing of breakpoint locations
between these nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves. 
Overall, Fig.~\ref{fig-inferred-args} shows that \tsinfer\ and \relate\ ARGs contain a level
of detail about recombination that lies somewhere between a sequence of unrelated local
trees on one extreme and an ARG with precisely specified recombination events on the other.

% Here we move on to coalescence
As for inferred coalescences, the shading (``coalescent span'')
reveals that many non-recombinant (green) nodes are
locally unary (see section \ref{sec-locally-unary-edges}). In other words,
in some regions of the genome these nodes are present in the ancestry without
corresponding to coalescence of ancestral material. As argued previously, this can provide useful
information about shared ancestry. Note however, that such unary segments are not inferred
in the \relate\ ARG: other than the samples it has entirely solid nodes.
Another clear difference involves coalescent nodes with more than two children (polytomies);
in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
uncertainty in the order of coalescent events, in an equivalent
manner to the imprecision associated with recombination nodes having more
than two parents. Only \tsinfer\ creates such polytomies, the clearest of which
involves the nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}. These
happen to have identical sequences, but because \kwarg, \argweaver, and \relate\ require
bifurcating trees, they are forced to pick an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
relationship between these samples). 
More generally, the number of nodes in Fig.~\ref{fig-inferred-args} C
that have more than two children hints at considerable uncertainty in
the order of branching elsewhere in the ARG.

@@

% TODO add additional explanation of the bottom row, if enough isn't written below
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
This illustrates that both \tsinfer\ and (to a lesser extent) \relate\ have edges
that span multiple tree boundaries, indicating that they are not inferring
a series of \emph{unrelated} local trees.
%(i.e., with either unlabelled internal
%nodes or differently labelled nodes in each tree).
For example, both infer that \noderef{Af-f}, \noderef{Fr-f} and \noderef{Wa-f}
share a common ancestor along
the entire sequence. However, in comparison to \kwarg\ and \argweaver\ neither
method results in extensive node sharing in the
oldest time periods. This primarily reflects the fact that they do not encode topology
above the local MRCAs, also indicated by multiple roots in the ARG
(nodes with zero parents, in blue), compared to the
single ARG root in the \kwarg\ and \argweaver\ genealogies. In terms of inference,
\tsinfer\ and \relate\ show a more pragmatic representation: we do not expect
genetic variation to be very informative about topological relationships
between nodes that that are older than the local MRCAs.
